---
title: "02.Amplicon_Analysis"
output: html_document
date: "2026-01-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Fierer Lab, University of Colorado Boulder
### Versions:
#### Version 1: October 2nd 2020 (presented by Cliff to CU Boulder CME)
#### Version 2: May 26, 2023 (presented by Nick in New Zealand)
#### Version 3: March 26, 2024 (shared to GitHub, shared with CU grad students)
### This tutorial will cover mctoolsR and some commonly used alpha and beta diversity statistics and plotting. You can do many of the same things with the phyloseq package. We use mctoolsR mainly because it was designed by a former student in the Fierer Lab (Jon Leff) and our dada2 pipeline is set up to export into the format required.

### This tutorial was written with R 4.2.3, mctoolsR 0.1.1.9, and dplyr 1.1.3.
### It has been tested and is working with R 4.2.3 and R 4.3.3

# Setup
## Note the code below uses the example dataset from mctoolsR (https://github.com/leffj/mctoolsr). We will note where you can change the inputs and code to use your own data instead

```{r}
# R packages required for this tutorial
suppressMessages(library(mctoolsr)) # For inputting, rarefying, filtering, taxonomic analyses
suppressMessages(library(plyr)) # For hulls
suppressMessages(library(tidyverse)) # For data manipulation and plotting
suppressMessages(library(RColorBrewer)) # For color palettes
suppressMessages(library(vegan)) # For alpha and beta diversity analyses
suppressMessages(library(indicspecies)) # For indicator species
suppressMessages(library(car)) # For Levene Test and ANOVA
suppressMessages(library(PMCMRplus)) # For Nemenyi posthoc test
suppressMessages(library(ggh4x)) # For nested facet plots
suppressMessages(library(emmeans)) # For pairwise comparisons
suppressMessages(library(multcomp)) # For automated significant difference letters
suppressMessages(library(zCompositions)) # CLR
suppressMessages(library(compositions)) # Aitchison
suppressMessages(library(reshape2)) # For melting
suppressMessages(library(cowplot)) # For plotting
suppressMessages(library(dplyr)) # For plotting

# Custom functions
source("cliffplot_taxa_bars.R") # https://github.com/cliffbueno/Alpha-Beta-Tutorial
source("plot_multipatt_asv.R") #https://github.com/cliffbueno/Alpha-Beta-Tutorial
```
# Loading Data
```{r}
# Loading ASV Table
tax_table_fp = "Fang_ASVTab_Filt.txt"

# Loading mapping file
map_fp = "Fang_Metadata.txt"


# Combine the ASV table and mapping file into an mctoolsR object
input = load_taxa_table(tax_table_fp, map_fp)
```

# Initial data examinination 
```{r}
# This is done by getting the column sums of the sequence table, and we'll sort it too.
sort(colSums(input$data_loaded))

# We could also save this as an object and plot it
seqcounts <- as.data.frame(sort(colSums(input$data_loaded)))

mean(sort(colSums(input$data_loaded))) #360.64
min(sort(colSums(input$data_loaded))) #1
max(sort(colSums(input$data_loaded))) #5893

seqcounts <- as.data.frame(sort(colSums(input$data_loaded))) %>%
  dplyr::rename("seqs" = "sort(colSums(input$data_loaded))") %>%
  rownames_to_column(var = "sampleID")

ggplot(seqcounts, aes(reorder(sampleID, seqs, mean), seqs)) + # Dataframe and variables
  geom_bar(stat = "identity") + 
  labs(y = "# Reads", x = "Sample") + 
  coord_flip() +
  theme_classic() + 
  theme(axis.text.y = element_text(size = 10))
```
# ALPHA DIVERSITY
```{r}
input$map_loaded$rich <- specnumber(input$data_loaded, 
                                             MARGIN = 2)
mean(input$map_loaded$rich) #9.28
min(input$map_loaded$rich) #1
max(input$map_loaded$rich) #46

input$map_loaded$shannon <- diversity(input$data_loaded, 
                                               index = "shannon", 
                                               MARGIN = 2)
mean(input$map_loaded$shannon) #1.3
min(input$map_loaded$shannon) #0
max(input$map_loaded$shannon) #3.3

write.csv(input$map_loaded,"Setup.csv")
```
## Rich
```{r}
# Two categories example
leveneTest(input$map_loaded$rich ~ input$map_loaded$Group)
# Variance homogeneous (p = 0.2618)

shapiro.test(input$map_loaded$rich)
# Richness not normally distributed (p =  3.513e-05)

wilcox.test(input$map_loaded$rich ~ input$map_loaded$Group)
# No sig diff

rich_comp<-ggplot(input$map_loaded, aes(Group, rich, colour = Group)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(size = 3, alpha = 0.5) +
  labs(x = "Group", y = "ASV Richness", colour = "Group") +
  scale_y_continuous(limits=c(0, 50), breaks = c(0,10,20,30,40,50))+
  scale_colour_manual(values = c("lightblue","darkgrey")) +
    ggtitle("p > 0.99")+
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14),
        legend.position = "none",
        axis.title.x = element_blank())
rich_comp
```

## shannon
```{r}
# Two categories example
leveneTest(input$map_loaded$shannon ~ input$map_loaded$Group)
# Variance homogeneous (p = 0.2618)

shapiro.test(input$map_loaded$shannon)
# Richness normally distributed (p =  0.3263)

wilcox.test(input$map_loaded$shannon ~ input$map_loaded$Group)
# No sig diff

Shannon_comp<-ggplot(input$map_loaded, aes(Group, shannon, colour = Group)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(size = 3, alpha = 0.5) +
  labs(x = "Group", y = "Shannon diversity", colour = "Group") +
  scale_y_continuous(limits=c(0, 4), breaks = c(0,1,2,3,4))+
  scale_colour_manual(values = c("lightblue","darkgrey")) +
    ggtitle("p = 0.50")+
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14),
        legend.position = "none",
        axis.title.x = element_blank())
Shannon_comp
```
## Seq
```{r}
library(ggbreak)

mean(input$map_loaded$Seq)

# Two categories example
leveneTest(input$map_loaded$Seq ~ input$map_loaded$Group)
# Variance homogeneous (p = 0.6033)

shapiro.test(input$map_loaded$Seq)
# Richness not normally distributed (p =  9.77e-10)

# If Levene's Test or Shapiro-Wilk Test p < 0.05, use Wilcoxon Test
wilcox.test(input$map_loaded$Seq ~ input$map_loaded$Group)
# No sig diff p= 0.7857

# Change colors to be colorblind friendly (e.g., red and blue is better than red and green)
Seq_comp<-ggplot(input$map_loaded, aes(Group, log(Seq), colour = Group)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(size = 3, alpha = 0.5) +
  labs(x = "Group", y = "log(Reads)", colour = "Group") +
  #coord_cartesian(ylim = c(0,800))+
  scale_y_log10()+ 
  # scale_y_break(c(800, 5800)) + 
  ggtitle("p = 0.79")+
  scale_y_continuous(limits=c(0, 10), breaks = c(0,2,4,6,8,10))+
  scale_colour_manual(values = c("lightblue","darkgrey")) +
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14),
        legend.position = "none",
        axis.title.x = element_blank())
Seq_comp
```
```{r}
FigS_RRS<-plot_grid(Seq_comp,rich_comp,Shannon_comp, nrow=1, ncol=3, align="vh", labels = "AUTO")

save_plot(plot = FigS_RRS, device="png", base_height= 3, base_width=9, filename = "2026_04_06-Supp_Fig_RRS.png")
```

# BETA DIVERSITY

## Bray-Curtis and NMDS
```{r}
# Calculate dissimilarity matrix
dm <- calc_dm(input$data_loaded)
# NOTE: calc_dm uses Bray-Curtis

# Let's calculate an ordination from the distance matrix
ord <- calc_ordination(dm, 'nmds')
# NOTE: You can also run pcoa and others by changing the code above from NMDS

# Now plot the ordination using the default mctoolsR function.
BD_NMDS<-plot_ordination(input, ord, 'Group', 'Group',hulls = TRUE) +
 scale_fill_manual(values =c("Ice" = "lightblue", "Tephra"="darkgrey")) +
   scale_color_manual(values =c("Ice" = "lightblue", "Tephra"="darkgrey"))+
  ggtitle("Bray-Curtis and NMDS")+
  theme_bw()+
  theme(legend.title = element_blank(), 
        legend.position = "right",
        axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14))

BD_NMDS
```

## Bray-Curtis and PCoA
```{r}
# Bray-Curtis dissimilarity matrix. Note: By default the data are square-root transformed.
bc <- calc_dm(input$data_loaded)

# Principle Coordinates Analysis (PCoA)
pcoa <- cmdscale(bc, k = nrow(input$map_loaded) - 1, eig = T)

# Variation Explained
eigenvals(pcoa)/sum(eigenvals(pcoa)) # 23.3, 16.6 % variation explained

# Make axis labels with % variation explained rounded to 1 digit.
pcoaA1 <- paste("PC1: ", round((eigenvals(pcoa)/sum(eigenvals(pcoa)))[1]*100, 1), "%")
pcoaA2 <- paste("PC2: ", round((eigenvals(pcoa)/sum(eigenvals(pcoa)))[2]*100, 1), "%") 

# Save Axis 1 and 2 scores to the mapping file
input$map_loaded$Axis01 <- scores(pcoa)[,1]
input$map_loaded$Axis02 <- scores(pcoa)[,2]

# Function for making a convex hull
find_hull <- function(df) df[chull(df$Axis01, df$Axis02),]

# Calculate hulls and save to dataframe
micro.hulls <- ddply(input$map_loaded, c("Group"), find_hull)

# Plot
BD_PCOA<-ggplot(input$map_loaded, aes(Axis01, Axis02, colour = Group, shape = Group)) +
  geom_polygon(data = micro.hulls, aes(colour = Group, fill = Group),
               alpha = 0.1, show.legend = F) +
  geom_point(size = 3, alpha = 0.5) +
    ggtitle("Bray-Curtis and PCoA")+
  labs(x = pcoaA1, 
       y = pcoaA2,
       colour = "Group",
       shape = "Group") +
 scale_fill_manual(values =c("Ice" = "lightblue", "Tephra"="darkgrey")) +
   scale_color_manual(values =c("Ice" = "lightblue", "Tephra"="darkgrey"))+
  theme_bw()+
  theme(legend.title = element_blank(), 
        legend.position = "right",
        axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14))

BD_PCOA
```

## Aitchison Matrix and PCA
```{r}
# First perform zero replacement and center log ratio (CLR) transformation. Use filtered but unrarefied data for this ("input_filt").
otu_czm <- cmultRepl(t(input$data_loaded), label = 0, method = "CZM")
otu_clr <- clr(otu_czm)
aclr <- compositions::dist(otu_clr)

# Filter dropped samples
# Columns and rows containing more than 80% zeros/unobserved values were deleted
# This is the bad thing about this method. We have now lost some samples and ASVs.
# You can also avoid this by adjusting the z.warning.
# Here we just proceed with the default.
dim(t(input$data_loaded)) # 28 samples, 933 ASVs
dim(otu_czm) # 23 samples, 144 ASVs

# Make a sampleID column
input$map_loaded$sampleID <- rownames(input$map_loaded)

# Filter out sampleIDs not in the CLR dataset. Save as new mctoolsR object.
input_clr <- filter_data(input,
                              filter_cat = "sampleID",
                              keep_vals = rownames(otu_czm))

# PCA (principle components analysis)
d.pcx <- prcomp(aclr)

# % variation explained
d.mvar <- sum(d.pcx$sdev^2)

# Make axes labels with % variation explained
PC1 <- paste("PC1: ", round((sum(d.pcx$sdev[1]^2)/d.mvar)*100, 1), "%")
PC2 <- paste("PC2: ", round((sum(d.pcx$sdev[2]^2)/d.mvar)*100, 1), "%")

# Save scores to map_loaded dataframe
input_clr$map_loaded$Axis01 <- d.pcx$x[,1]
input_clr$map_loaded$Axis02 <- d.pcx$x[,2]

# Transform the scores
input_clr$map_loaded$Axis01 <- input_clr$map_loaded$Axis01/sqrt(sum((input_clr$map_loaded$Axis01 - mean(input_clr$map_loaded$Axis01))^2))
input_clr$map_loaded$Axis02 <- input_clr$map_loaded$Axis02/sqrt(sum((input_clr$map_loaded$Axis02 - mean(input_clr$map_loaded$Axis02))^2))

# Calculate hulls
micro.hulls <- ddply(input_clr$map_loaded, c("Group"), find_hull)

# Plot
DB_PCA<-ggplot(input_clr$map_loaded, aes(Axis01, Axis02, colour = Group, shape = Group)) +
  geom_polygon(data = micro.hulls, aes(colour = Group, fill = Group),
               alpha = 0.1, show.legend = F) +
  geom_point(size = 3, alpha = 0.5) +
  ggtitle("Aitchison Matrix and PCA")+
  labs(x = PC1, 
       y = PC2,
       colour = "Group",
       shape = "Group") +
  scale_fill_manual(values =c("Ice" = "lightblue", "Tephra"="darkgrey")) +
   scale_color_manual(values =c("Ice" = "lightblue", "Tephra"="darkgrey"))+
  theme_bw()+
  theme(legend.title = element_blank(), 
        legend.position = "right",
        axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14))

DB_PCA
# Note: This is very consistent with the Bray-Curtis and PCoA.
```
```{r}
FigS_BetaDiv<-plot_grid(BD_NMDS,BD_PCOA,DB_PCA, nrow=3, ncol=1, align="vh", labels = "AUTO")
save_plot(plot = FigS_BetaDiv, device="png", base_height= 9, base_width=6, filename = "2026_04_06-Supp_Fig_BetaDiv.png")
```

## PERMANOVA, PERMDISP
```{r}
# PERMANOVA (multivariate version of ANOVA)
JC <- calc_dm(input$data_loaded, method="jaccard")
m2 <- adonis2(JC ~ input$map_loaded$Group, permutations = 999)

set.seed(1223) # To make reproducible 
m <- adonis2(bc ~ input$map_loaded$Group, permutations = 999)
#Not_Sig_effect(0.823)

# What about pairwise PERMANOVA? This is similar to a Tukey posthoc an an ANOVA.
# Grouping variable must be a factor
input$map_loaded$Group <- as.factor(input$map_loaded$Group)
set.seed(1223)
m_pw <- calc_pairwise_permanovas(bc, input$map_loaded, "Group")
m_pw # Each group is significantly different from the other.

# PERMDISP (multivariate version of Levene's Test)
m1 <- betadisper(bc, input$map_loaded$Group)
anova(m1) # Dispersion different
#5.153e-05

Plot1<-plot(m1,col = c("lightblue","darkgrey"), ellipse = TRUE, hull = FALSE)
Plot2<-boxplot(m1,col = c("lightblue","darkgrey"))

##SO No DIfference in average, difference in dispersion
#PERMANOVA tests if groups differ in position (centroids) or spread (dispersion).
#PERMDISP (betadisper): Specifically tests the homogeneity of multivariate dispersion (Are groups equally spread out?).
```

## Hierarchical Clustering
### Another option to examine community composition among your samples and treatments is to see how they cluster. You can already see how they cluster in the ordination plots, but this is another method that can be used.
```{r}
h <- hclust(bc)
plot(h)
```

# TAXONOMIC ANALYSIS
## Indicator Taxa
### Let's first check for indicator taxa for our grouping variables. We will try two methods, SIMPER and MULTIPATT. SIMPER will list how much each ASV contributes to dissimilarity among groups. MULTIPATT will list indicators for each individual group as well as combinations of groups. These methods use a large amount of computer memory and can sometimes take a while or max out your computer, depending on the dataset size. It will run faster/better on a server or if you just run it on the most abundant taxa instead of tens of thousands of ASVs. In this case, this is a simple dataset and we won't have any problems.
```{r}
# SIMPER (list how much each ASV contributes to dissimilarity among groups)
sim <- simper(t(input$data_loaded), 
              input$map_loaded$Group)
s <- summary(sim)
# Let's look at the top 5 contributing to dissimilarity between mushrooms and strawberries/spinach
Out145<-head(s$Tephra_Ice, n = 145)

write.csv(Out145, "2026_04_06-SIMPER_ExportALL.csv")
# in this case it looks like the top 5 ASVs are expalining a lot of the difference (~40%)
```
```{r}
# Loading ASV Table
tax_table_relabun = "/Users/nbd/Desktop/Work/Research/Antarctic_Cores/03_Amplicon/02.Filt_ASVTab/2026_01_02-Fang_ASVTab_RELABUN.txt"
# Loading mapping file
# Combine the ASV table and mapping file into an mctoolsR object
input_RA = load_taxa_table(tax_table_relabun, map_fp)

input_RA_filt<-filter_data(input = input_RA, filter_cat = "sampleID", filter_vals = c("Fang11",  "Fang4", "Fang13", "Fang16", "Fang12", "Fang23", "Fang10",  "Fang9"))

# MULTIPATT (list ASVs associated with each group)
# Groups can be individual treatments or groups of treatments
set.seed(1223) # For reproducability
mp <- multipatt(t(input_RA$data_loaded), 
                input_RA$map_loaded$Group, 
                func = "IndVal", 
                control = how(nperm=999))
summary(mp)

# Now lets use a custom function for plotting just 1 group indicators and choosing p value and indicator value cutoffs.
# Rerun the multipatt with the r.g func.
set.seed(1223) # For reproducibility
mp <- multipatt(t(input_RA$data_loaded), 
                input_RA$map_loaded$Group, 
                func = "r.g", 
                control = how(nperm=999))
summary(mp)

# This function was built with an 8th taxonomy column containing the ASV ID. Make here.
input_filt_rar$taxonomy_loaded$taxonomy8 <- rownames(input_filt_rar$taxonomy_loaded)

# It takes a summarized taxonomy dataframe, made like this.
tax_sum_asv <- summarize_taxonomy(input_filt_rar, level = 8, report_higher_tax = FALSE)

# The group must be a factor
input_filt_rar$map_loaded$Sample_type <- as.factor(input_filt_rar$map_loaded$Sample_type) 

# Run the function.
plot_multipatt_asv(mp_obj = mp, 
                   input = input_filt_rar,
                   tax_sum = tax_sum_asv,
                   group = "Sample_type",
                   filter = FALSE,
                   abund = "% Rel. Abund.",
                   qcut = 0.05, # Adjusted p value< 0.05
                   rcut = 0.5) # Can change r cutoff to be more or less selective

# This will plot the ASVs that passed the cutoffs. It will plot how correlated they are with a group as well as their % relative abundance. It will state the phylum and ASV ID.
```

## OTU/ASV Specific functions
```{r}
# Let's look at the top taxa across samples (Can change the number you want seen)
return_top_taxa(input = input, number_taxa = 10)
# Calculates the top taxa based on number of counts - taking the row means for each OTU

# What taxa are common? Give them rarefied input, then level = "what level we want to summarize our information for". Check input_filt_rar$taxonomy_loaded for the levels. level 2 is phylum level
tax_sum_phyla <- summarize_taxonomy(input, level = 2, report_higher_tax = FALSE)
#level taxa=false- just look at that level, not at whole taxonomic classification

# How many phyla can we detect in this dataset
nrow(tax_sum_phyla)

# Check the most abundant phyla across dataset
tax_sum_phyla_Means <- sort(rowMeans(tax_sum_phyla), decreasing = T)

# This is now sorted from most abundant to least abundant, based on means across all samples
# Check the top 5 phyla
tax_sum_phyla_Means[1:5]

# Now lets try summarizing at the family level (level 5)
tax_sum_families <- summarize_taxonomy(input, level = 5, report_higher_tax = FALSE)
nrow(tax_sum_families)
tax_sum_families_Means <- sort(rowMeans(tax_sum_families), decreasing = T)
tax_sum_families_Means[1:5]

```

## Other useful plotting and analysis functions
### Base mctoolsR has some other nice functions for running statistical tests on taxonomic abundances as well as plotting heatmaps and stacked bar plots. We also demonstrate some customizations on the base functions. For more information on mctoolsR functions, see the tutorial at https://github.com/leffj/mctoolsr.
```{r}
# Summarize at family level. Choose different levels with the level argument
tax_sum_families <- summarize_taxonomy(input, level = 5, report_higher_tax = FALSE)

# Default heatmap
plot_ts_heatmap(tax_sum_families, 
                input$map_loaded, 
                0.01, 
                'Description')

# Add some customization
plot_ts_heatmap(tax_sum_families, 
                input$map_loaded, 
                0.01, 
                'Description',
                rev_taxa = T) +
  coord_flip() +
  theme(axis.text.x = element_text(size = 12, angle = 45, vjust = 1))

```
```{r}
# Default (choose how many taxa you want with num_taxa)
Family_bars<-plot_taxa_bars(tax_sum_families,
               input$map_loaded,
               "Description",
               num_taxa = 20, data_only = TRUE)

Phyla_bars<-plot_taxa_bars(tax_sum_phyla,
               input$map_loaded,
               "Description",
               num_taxa = 12, data_only = TRUE)

#write_csv(Family_bars, "2026_01_05-Family_Bars_Export.csv")
#write_csv(Phyla_bars, "2026_01_05-Phyla_bars_Export.csv")

```
```{r}
# Add some customization
plot_taxa_bars(tax_sum_families,
               input$map_loaded,
               "Description",
               num_taxa = 10) +
  labs(x = "Sample", y = "Relative Abundance", fill = "Family") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14),
        axis.text.x = element_text(angle=45))

# Add even more customization. plot_taxa_bars has a data_only argument. So you can save the dataset created and then write your own plot!
bars <- plot_taxa_bars(tax_sum_families,
                       input_filt_rar$map_loaded,
                       "Sample_type",
                       num_taxa = 10,
                       data_only = TRUE)
ggplot(bars, aes(group_by, mean_value, fill = taxon)) +
  geom_bar(stat = "identity", colour = "black", linewidth = 0.25) +
  labs(x = "Sample Type", y = "Relative Abundance", fill = "Family") +
  scale_fill_brewer(palette = "Paired") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14))

# Different style - classic theme, no black lines between taxa.
ggplot(bars, aes(group_by, mean_value, fill = taxon)) +
  geom_bar(stat = "identity", linewidth = 0.25) +
  labs(x = "Sample Type", y = "Relative Abundance", fill = "Family") +
  scale_fill_brewer(palette = "Paired") +
  theme_classic() +
  theme(axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14))
```
```{r}

PlotK<-cliffplot_taxa_bars(input = input, level = 1, variable = "Group")
PlotP<-cliffplot_taxa_bars(input = input, level = 2, variable = "Group")
PlotC<-cliffplot_taxa_bars(input = input, level = 3, variable = "Group")
PlotO<-cliffplot_taxa_bars(input = input, level = 4, variable = "Group")
PlotF<-cliffplot_taxa_bars(input = input, level = 5, variable = "Group")
PlotG<-cliffplot_taxa_bars(input = input, level = 6, variable = "Group")

Supp_Group_TaxBars<-plot_grid(PlotK,PlotP,PlotC,PlotO,PlotF,PlotG, nrow=3,ncol=2, align="vh", labels = "AUTO")
Supp_Group_TaxBars

#save_plot(plot = Supp_Group_TaxBars, device="png", base_height= 12, base_width=10, filename = "2026_01_05-Supp_Fig_TaxBar_Group.png")
```

```{r}
bars_labels<-read_csv("/Users/nbd/Desktop/Work/Research/Antarctic_Cores/AA.Plotting/Figs_Taxa_Bars/Label_plot.csv")

Stacked_Labels<-ggplot()+
  geom_bar(stat="identity", data= bars_labels, aes(x = fct_reorder(groupby, OrderF, .desc = FALSE), y= mean_value, fill= Color2))+
   scale_fill_manual(values = c("Tephra" = "darkgrey","Ice" = "lightblue"))+
  coord_flip()+
    theme_bw() +
  theme(
    axis.text.x = element_blank(),
    #legend.position = "none",
    axis.title.x =element_blank(),
    axis.text.y = element_text(size = 10),
    axis.ticks.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.text = element_text(size=10),
    legend.title = element_text(face = "bold", size=10))+
  labs(x = "Site", y = "Proportion of Reads",fill= "Sample type")

Stacked_Labels

#save_plot(plot = Stacked_Labels, device="png", base_height= 6, base_width=14, filename = "2026_01_05-Fang_Family_Stacked_labelsexport.png")

```

```{r}
Fam_bars_Plot<-read_csv("/Users/nbd/Desktop/Work/Research/Antarctic_Cores/AA.Plotting/Figs_Taxa_Bars/2026_01_05-Family_bars_Form.csv")

mycolors<-c("lightblue",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"lightblue",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"lightblue",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"lightblue",
"darkgrey",
"lightblue",
"darkgrey")

Stacked_family<-ggplot()+
  geom_bar(stat="identity", data= Fam_bars_Plot, aes(x = fct_reorder(group_by, OrderF, .desc = FALSE), y= mean_value, fill= taxon))+
   #scale_fill_manual(values = c("Z_Other" = "#332288","Verrucomicrobiota" = "deepskyblue4","Proteobacteria" = "#0077BB","Planctomycetota" = "#88CCEE","Firmicutes" = "darkseagreen2","Deinococcota" = "#44AA99","Cyanobacteria" = "#117733","Chloroflexota" = "#355f8d","Bacteroidota" = "#999933","Actinobacteriota" = "#CC6677","Acidobacteriota" = "#AA4499"))+
  coord_flip()+
    theme_bw() +
  theme(
    axis.text.x = element_text(size = 10, hjust=1),
    axis.title.x = element_text( face = "bold",size = 10),
    axis.text.y = element_blank(),
    axis.ticks.y  = element_blank(),
    axis.title.y=element_blank(),
    legend.text = element_text(size=10),
    legend.title = element_text(face = "bold", size=10))+
  labs(x = "Site", y = "Proportion of Reads",fill= "Phyla")

Stacked_family

FamilyBars_Plot<-plot_grid(Stacked_Labels, Stacked_family, align = "vh",rel_widths = c(3,6))


#save_plot(plot = FamilyBars_Plot, device="png", base_height= 6, base_width=14, filename = "2026_01_05-Fang_Family_Stacked_Barexport.png")
```

```{r}
phyla_bars_Plot<-read_csv("/Users/nbd/Desktop/Work/Research/Antarctic_Cores/AA.Plotting/Figs_Taxa_Bars/2026_01_05-Phyla_bars_Form.csv")

mycolors<-c("lightblue",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"lightblue",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"lightblue",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"darkgrey",
"lightblue",
"darkgrey",
"lightblue",
"darkgrey")

Stacked_Phyla<-ggplot()+
  geom_bar(stat="identity", data= phyla_bars_Plot, aes(x = fct_reorder(group_by, OrderF, .desc = FALSE), y= mean_value, fill= taxon))+
   scale_fill_manual(values = c("Other"="grey","Verrucomicrobiota" = "#332288","Thermoproteota" = "deepskyblue4","Pseudomonadota" = "#0077BB","Planctomycetota" = "#88CCEE","Deinococcota" = "darkseagreen2","Cyanobacteriota" = "#117733","Chloroflexota" = "#44AA99","Bdellovibrionota" = "#355f8d","Bacteroidota" = "#999933","Bacillota" = "#CC6677","Actinomycetota" = "#AA4499", "Acidobacteriota" = "mediumpurple3"))+
  coord_flip()+
    theme_bw() +
  theme(
    axis.text.x = element_text(size = 10, hjust=1),
    axis.title.x = element_text( face = "bold",size = 10),
    axis.text.y = element_blank(),
    axis.ticks.y  = element_blank(),
    axis.title.y=element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size=10),
    legend.title = element_text(face = "bold", size=10))+
  labs(x = "Site", y = "Proportion of Reads",fill= "Phyla")


Stacked_Phyla
```
```{r}
Fig1<-plot_grid(Fang_Core_plot,Stacked_Labels, Stacked_Phyla,Samples_Rich, Samples_Shan,nrow=1, align = "vh",rel_widths = c(3,2,6,3,3), labels = c("A","B","","C","D"))
#save_plot(plot = Fig1, device="png", base_height= 6, base_width=27, filename = "2026_01_05-Fig1_combined.png")
```
```{r}
Samples_Reads<-ggplot(data= bars_labels, aes(x = fct_reorder(groupby, OrderF, .desc = FALSE), y= log(Seqs), fill= Color2))+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=1)+
  #scale_x_discrete(position = "top")+
    #scale_y_continuous(limits = c(0, 4.5), breaks = c(0,1.5,3,4.5)) +
  coord_flip()+
    theme_bw() +
  theme(
    axis.text.x = element_text(size = 10, hjust=1),
    axis.title.x = element_text( face = "bold",size = 10),
    axis.text.y = element_blank(),
    axis.ticks.y  = element_blank(),
    axis.title.y=element_blank(),
    legend.text = element_text(size=10),
    legend.title = element_text(face = "bold", size=10))+
  labs(x = "Site", y = "Proportion of Reads",fill= "Phyla")

Samples_Reads
```
```{r}
Samples_Rich<-ggplot(data= bars_labels, aes(x = fct_reorder(groupby, OrderF, .desc = FALSE), y= Rich, fill= Color2))+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=3.0)+
  #scale_x_discrete(position = "top")+
    #scale_y_continuous(limits = c(0, 4.5), breaks = c(0,1.5,3,4.5)) +
  coord_flip()+
    scale_fill_manual(values =c("Ice" = "lightblue", "Tephra"="darkgrey")) +
    theme_bw() +
  theme(
    axis.text.x = element_text(size = 10, hjust=1),
    axis.title.x = element_text( face = "bold",size = 10),
    axis.text.y = element_blank(),
    #axis.ticks.y  = element_blank(),
    axis.title.y=element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    #legend.text = element_text(size=10),
    legend.title = element_blank())+
  labs(x = "Site", y = "Richness",fill= "Phyla")

Samples_Rich
```
```{r}
Samples_Shan<-ggplot(data= bars_labels, aes(x = fct_reorder(groupby, OrderF, .desc = FALSE), y= Shannon, fill= Color2))+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=3.0)+
  #scale_x_discrete(position = "top")+
    scale_y_continuous(limits = c(0, 4), breaks = c(0,1,2,3,4)) +
  coord_flip()+
    scale_fill_manual(values =c("Ice" = "lightblue", "Tephra"="darkgrey")) +
    theme_bw() +
  theme(
    axis.text.x = element_text(size = 10, hjust=1),
    axis.title.x = element_text( face = "bold",size = 10),
    axis.text.y = element_blank(),
    #axis.ticks.y  = element_blank(),
    axis.title.y=element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    #legend.text = element_text(size=10),
    legend.title = element_blank())+
  labs(x = "Site", y = "Richness",fill= "Phyla")

Samples_Shan
```
```{r}
# Make sampleID column
input_filt_rar$map_loaded$sampleID <- rownames(input_filt_rar$map_loaded)

# Summarize taxonomy. Let's do families.
tax_sum_families <- summarize_taxonomy(input_filt_rar, level = 5, report_higher_tax = FALSE)

# Get the plotting data. We'll also join the rest of map_loaded.
bars <- plot_taxa_bars(tax_sum_families,
                       input_filt_rar$map_loaded,
                       "sampleID",
                       num_taxa = 12,
                       data_only = TRUE) %>%
  mutate(taxon = fct_rev(taxon)) %>%
  left_join(., input_filt_rar$map_loaded, by = c("group_by" = "sampleID"))

# Run stats and add an asterisk to their name in the legend if significantly affected by treatment
fam_stats <- taxa_summary_by_sample_type(tax_sum_families, 
                                         input_filt_rar$map_loaded, 
                                         type_header = 'Sample_type', 
                                         filter_level = 0.01, 
                                         test_type = 'KW') %>%
  filter(rownames(.) %in% bars$taxon) %>%
  arrange(desc(rownames(.))) %>%
  mutate(Sig = ifelse(pvalsFDR < 0.05, "Pfdr < 0.05", "Pfdr > 0.05")) %>%
  mutate(Star = ifelse(pvalsFDR < 0.05, "*", "")) %>%
  rownames_to_column(var = "taxon") %>%
  arrange(match(taxon, levels(bars$taxon))) %>%
  mutate(StarLab = paste(taxon, Star, sep = " "))

# Plot
ggplot(bars, aes(group_by, mean_value, fill = taxon)) +
  geom_bar(stat = "identity", linewidth = 0.25) +
  labs(x = "Sample Type", y = "Relative Abundance", fill = "Family") +
  scale_fill_manual(values = c("grey90", brewer.pal(12, "Paired")[12:1]),
                    labels = c("Other", fam_stats$StarLab)) +
  scale_y_continuous(expand = c(0.01, 0.01)) + 
  facet_nested(~ Farm_type + Sample_type, space = "free", scales = "free_x") +
  theme_classic() +
  theme(axis.title.y = element_text(face = "bold", size = 12),
        axis.title.x = element_blank(),
        axis.text.y = element_text(size = 10),
        axis.text.x = element_text(size = 8, angle = 90, hjust = 1, vjust = 0.5),
        axis.ticks.x = element_blank(),
        axis.line.x = element_blank(),
        strip.text = element_text(size = 6),
        strip.background = element_rect(linewidth = 0.2),
        axis.line.y = element_blank(),
        legend.margin = margin(0, 0, 0, 0, unit = "pt"),
        legend.box.margin = margin(0, 0, 0, 0, unit = "pt"),
        legend.key.size = unit(0.3, "cm"),
        panel.spacing.x = unit(c(0.2, 0.2, 0.4, 0.2, 0.2), "lines"))

# Test (run a Kruskal-Wallis test on all families with mean rel abund. > filter_level in at least one of the factor levels)
taxa_summary_by_sample_type(tax_sum_families, 
                            input_filt_rar$map_loaded, 
                            type_header = 'Sample_type', 
                            filter_level = 0.05, 
                            test_type = 'KW')
# Or Wilcoxon for 2 levels (farm type)
taxa_summary_by_sample_type(tax_sum_families, 
                            input_filt_rar$map_loaded, 
                            type_header = 'Farm_type', 
                            filter_level = 0.05, 
                            test_type = 'MW')
```

#Associated plot
```{r}
associated_ASVs<-read.csv("/Users/nbd/Desktop/Work/Research/Antarctic_Cores/03_Amplicon/04.Simper_out/2026_04_06-SIMPER_PLOT.csv")

label_colors<-associated_ASVs$Lab_Col_3

tephra_ice_Associated<-ggplot(associated_ASVs,
       aes(x = Group, 
           y = fct_reorder(Genus, Order, .desc = TRUE),
           colour = Group)) +
  geom_point(aes(size = Avg_Abund)) +
  scale_size(limits = c(0.001,300),range = c(1,7), name = waiver()) +
   # scale_size_continuous(range = c(0, 15)) + # Adjust as required.
   # scale_size_area(max_size = 10,name = waiver()) + # Adjust as required.
  scale_x_discrete(position = "top") +
  scale_color_manual(values = c("Tephra" = "darkgrey","Ice" = "lightblue"))+
  theme_classic()+
  labs(x = NULL, y = NULL, fill= NULL, size= "Average abundance (reads)") +
theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
      axis.text.y = element_text(face = "italic"))#color = label_colors))
        #panel.background = element_blank(),
        #panel.grid = element_blank())
  #      axis.ticks = element_blank())
#save_plot(plot = tephra_ice_Associated, device="png", base_height= 6, base_width=8, filename = "2026_04_07-Supp_Fig_Associated_taxa.png")
```

